---
title: "CDS 101 – Final Project Report"

author: "Team 3: Athir, Lorenzo, Abel"

author: "Group Name / Members"

date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "gmu-cds101.css"
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
---

> Replace this template text with your own writing, keeping the headings and overall structure aligned with the **CDS 101 project rubric**.

# 1. Problem Definition

<<<<<<< HEAD
Burglary is a crime that is the cause of significant safety and financial concerns for residents of urban areas. Understanding when burglaries are most likely to occur can help our communities, law enforcement, and policymakers better understand and implement preventative measures. The Washington Dc area is a large and densely populated area, making it an important case study for examining certain crimes and patterns of those crimes. 

The main research question for this project is: **Does the frequency of burglary incidents in the DC Metro area differ across different times of day?** The main objective of this project is to use historical crime data to examine how burglary occurrences vary by time of day and to test whether these differences are significant. We assume that reported burglary data accurately shows real crime patterns, and that the time of day classifications are recorded in a consistent way.

To address this research question, we will first load and inspect the DC Metro Crime dataset. We will then clean the data by filtering for burglary offenses and handling missing values in the time of day variable. Next, we will do some exploratory data analysis using summaries and visualizations to examine how burglary frequency varies by shift. Finally, we will perform a formal hypothesis test to determine whether the observed differences across times of day are statistically significant, and interpret the results.

=======
- Clearly state the main problem or question you are investigating.  
- Explain why it is interesting or important.  
- Define the objective(s) of your analysis.  
- Mention any key assumptions.
>>>>>>> c25410689d42c185141587df3b9feaaa8b98bf4b

# 2. Data Acquisition & Description

- Name and source of the dataset(s).  
- How you obtained the data (download link, API, etc.).  
- Brief description of the variables / features (with units if relevant).  
- Approximate size (rows, columns).  
- Any ethical, privacy, or bias considerations (if applicable).

```{r setup-data, echo=FALSE, message=FALSE, warning=FALSE}
# Example: load a dataset (replace with your own)
# library(readr)
# data <- read_csv("data/your_data.csv")
```

# 3. Data Cleaning & Preprocessing

Describe the steps you took to make the data usable, for example:

- Handling missing values (drop, impute, etc.).  
- Fixing types (dates, numeric vs. categorical).  
- Removing outliers (if done).  
- Creating new features (feature engineering).  
- Any scaling or encoding.

```{r cleaning, eval=FALSE}
# Example pseudo-code for cleaning:
# data_clean <- data |>
#   dplyr::filter(!is.na(target_variable)) |>
#   dplyr::mutate(
#     new_feature = some_transformation(old_feature)
#   )
```

# 4. Exploratory Data Analysis (EDA)

This section maps directly to the **EDA** part of the rubric:

- Summary statistics  
- Visualizations  
- Interpretation connected to the research question  

```{r eda-summary, message=FALSE, warning=FALSE}
# library(dplyr)
# summary(data_clean)
```

```{r eda-plot-example, message=FALSE, warning=FALSE, fig.cap="Example histogram of a numeric variable (EDA rubric)."}
# library(ggplot2)
# ggplot(data_clean, aes(x = some_numeric_variable)) +
#   geom_histogram(binwidth = 5, color = "white") +
#   labs(x = "Some Variable", y = "Count", title = "Distribution of Some Variable")
```

# 5. Visualization Quality and Storytelling

Use this section to satisfy the **Visualization Quality** rubric criterion:

- Explain why your plot types are appropriate.  
- Comment on labels, legends, colors, and overall readability.  
- Mention any steps you took to make plots accessible and interpretable.

# 6. Modeling Approach

Explain how you framed the problem and which models you chose:

- Type of task (regression, classification, etc.).  
- Baseline model or heuristic, if used.  
- Main model(s) chosen and why they are appropriate.

# 7. Model Implementation & Evaluation

This corresponds to the **Model Implementation & Evaluation** rubric criterion.

Describe:

- Features used.  
- Data splitting strategy (train/test or cross-validation).  
- Metrics used (accuracy, F1, RMSE, etc.).  
- Tables/plots summarizing performance.  
- A short interpretation of each metric/plot.

```{r modeling, eval=FALSE}
# Example structure:
# set.seed(123)
# train_index <- sample(seq_len(nrow(data_clean)), size = 0.8 * nrow(data_clean))
# train <- data_clean[train_index, ]
# test  <- data_clean[-train_index, ]
#
# model <- glm(target ~ ., data = train, family = binomial)
# preds <- predict(model, newdata = test, type = "response")
# # compute metrics...
```

# 8. Conclusions & Recommendations

Summarize the key takeaways:

- Answer your original research question(s) directly.  
- Highlight the most important patterns or relationships you found.  
- Discuss limitations (data size, bias, missing variables, etc.).  
- Suggest possible extensions or future work.

# 9. Code Quality & Reproducibility

Briefly document how someone else can reproduce your results:

- Which R scripts or Rmd files to run.  
- Any required R packages.  

```{r session-info, echo=FALSE}
sessionInfo()
```

# 10. References

List any references you used, such as:

- Dataset documentation.  
- Research papers or articles.  
- Tutorials or blog posts.

# Appendix (Optional)

Include extra plots, diagnostic checks, or model comparisons if needed.
