---
title: "CDS 101 – Final Project Report"

author: "Team 3: Athir, Lorenzo, Abel"


date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "gmu-cds101.css"
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
---

# 1. Problem Definition


Burglary is a crime that is the cause of significant safety and financial concerns for residents of urban areas. Understanding when burglaries are most likely to occur can help our communities, law enforcement, and policymakers better understand and implement preventative measures. The Washington Dc area is a large and densely populated area, making it an important case study for examining certain crimes and patterns of those crimes. 

The main research question for this project is: **Does the frequency of burglary incidents in the DC Metro area differ across different times of day?** The main objective of this project is to use historical crime data to examine how burglary occurrences vary by time of day and to test whether these differences are significant. We assume that reported burglary data accurately shows real crime patterns, and that the time of day classifications are recorded in a consistent way.

To address this research question, we will first load and inspect the DC Metro Crime dataset. We will then clean the data by filtering for burglary offenses and handling missing values in the time of day variable. Next, we will do some exploratory data analysis using summaries and visualizations to examine how burglary frequency varies by shift. Finally, we will perform a formal hypothesis test to determine whether the observed differences across times of day are statistically significant, and interpret the results.


# 2. Data Acquisition & Description

The dataset used for this project is the **DC Metro Crime Dataset**, obtained from Kaggle. It contains records of reported criminal incidents in the Washington, DC metro area.

The dataset was obtained by downloading the CSV file from the Kaggle website and storing it locally in the project data folder for analysis.

Key variables used in this project include:

- **offense (categorical)**: the type of crime committed (for example, burglary).

- **shift (categorical)**: the time of day the crime occurred, categorized as Day, Evening, or Midnight.

The full dataset contains 342,864 observations and 32 variables. Due to GitHub file size limits, the dataset was split into two CSV files, dc_crime_part1.csv and dc_crime_part2.csv, which together is the complete dataset.


This dataset is based on publicly released government crime reports, so it does not contain personal identifying information. However, we believe potential reporting bias may exist, as not all crimes are reported equally across neighborhoods or times. Additionally, policing practices and data collection methods may influence recorded crime frequencies.



```{r setup-data, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)

# We will load both parts of the dataset
dc_part1 <- read.csv("data/dc_crime_part1.csv")
dc_part2 <- read.csv("data/dc_crime_part2.csv")

#combine
dc_crime <- bind_rows(dc_part1,dc_part2)
```

# 3. Data Cleaning & Preprocessing

Before conducting our analysis, the raw DC Metro Crime dataset was cleaned and preprocessed to make sure everything was accurate and consistent.

Records with missing or not valid values in key variables, mainly shift (time of day) and offense, were removed from the dataset. These variables are required for answering the research question, so incomplete values could not be used.

Several variables were converted to appropriate data types:

- **shift** and **offense** were converted to categorical variables.

- **date** and **time-related fields** were formatted into proper date types where needed.
This made sure that grouping, filtering, and tests could be done correctly.

No numerical outliers were removed because the analysis focuses on categorical variables of burglary incidents and not continuous numeric measurements. Every valid burglary report was kept to keep the validity of the crime distribution.

The dataset was filtered to include only burglary offenses, since burglary is the main focus of the research question. This created a new working dataset we used for modeling and hypothesis testing. Additionally, the shift variable was used as the main grouping variable to compare burglary frequency across times of day.

No numeric scaling was required because the analysis does not rely on continuous predictors.


```{r cleaning}
library(tidyverse)

#Need to keep only the variables we need
dc_clean <- dc_crime %>%
  select(OFFENSE,SHIFT, REPORT_DAT)

#Need to remove missing values in the variables
dc_clean <- dc_clean %>%
  filter(!is.na(OFFENSE), !is.na(SHIFT))

#Now we need to convert the data types
dc_clean <- dc_clean %>%
  mutate(
    OFFENSE = as.factor(OFFENSE),
    SHIFT = as.factor(SHIFT),
    REPORT_DAT = as.Date(REPORT_DAT, format = "%m/%d/%Y")
  )

#Now we filter for Burglary only
dc_burglary <- dc_clean %>%
  filter(OFFENSE == "BURGLARY")
```

# 4. Exploratory Data Analysis (EDA)

This section presents descriptive statistics and visualizations of burglary incidents across different times of day. The goal is to explore whether burglary frequency varies by shift, which directly supports the project’s research question. 

```{r eda-summary, message=FALSE, warning=FALSE}
library(tidyverse)

burglary_summary <- dc_burglary %>%
  group_by(SHIFT) %>%
  summarize(
    Count = n(),
    Percentage = round(100 * n() / nrow(dc_burglary), 1)
  )

burglary_summary  
```

### **Interpretation**

The percentage column represents the share of total burglary incidents occurring in each shift. For example, the evening shift accounts for approximately 43.6% of all burglaries, followed by the day shift at 40% and the midnight shift at 16.4%.





```{r eda-plot-example, message=FALSE, warning=FALSE, fig.cap="Histogram showing the distribution of burglary incidents over time in the DC Metro area"}
fig.cap = "Histogram showing the distribution of burglary incidents over time in the DC Metro area"

ggplot(dc_burglary, aes(x = REPORT_DAT)) +
  geom_histogram(binwidth = 30) +
  labs(
    title = "Distribution of Burglary Incidents Over Time",
    x = "Report Date",
    y = "Number of Burglary Incidents"
  )
```

### **Interpretation**

This histogram shows how reported burglary incidents are distributed over time. The overall spread of the distribution indicates that burglary occurs consistently throughout the study period, with some variation in frequency across years and months. This confirms that the dataset captures long-term burglary patterns rather than being concentrated in a short time window.

# 5. Visualization Quality and Storytelling

```{r}
library(tidyverse)

dc_burglary %>%
  ggplot() +
    geom_bar(mapping = aes(x = SHIFT)) +
    labs(
    title = "Burglary Incidents by Shift",
    x = "Time of Day (Shift)",
    y = "Number of Burglary Incidents"
  )
```


```{r}
library(tidyverse)

#Extracts year from report date in dc_burglary data set
dc_burglary_year <- dc_burglary %>%
  mutate(Year = year(REPORT_DAT))

#Counts Burglaries by Shift and Year
heatmap_burglary <- dc_burglary_year %>%
  group_by(SHIFT, Year) %>%
  summarise(Count = n())

heatmap_burglary%>%
  ggplot()+
  geom_tile(mapping = aes(x = SHIFT, y = Year, fill = Count))+
  scale_fill_gradient(low = "lightblue", high = "red")+
  labs(
    title = "Heatmap of Burglary Incidents by Shift",
    x = "Time of Day (Shift)",
    y = "Year"
  )
  
```
We used a bar graph to clearly show differences in time and the number of burglaries. Additionally, we also used a heat map plot for our analysis to help with storytelling and showcase the areas of density when it comes to the number of burglaries during a specific time in a specific year. This allows us to showcase the consistency in the number of burglaries during a time of day (Shift), as well as any potential inconsistencies. 

These plots are appropriate because our analysis focuses on categorical variables. Since our focus is on the time of day burglaries occur the most.

Both plots have clear titles and labeled axes. Our bar graph used the default colors so that all users can clearly see the difference in the number of burglaries and time without being distracted by extra details. Additionally, the heatmap utilized a legend to showcase the corresponding color with the count. For example, light blue represents a lower amount of burglaries, and red represents a higher amount of burglaries.

Lastly, all graphs go from left to right in an ascending order, day -> evening -> midnight. Allowing readers to clearly visualize the amount of burglaries occurring based on time.

# 6. Modeling Approach

Explain how you framed the problem and which models you chose:

- Type of task (regression, classification, etc.).  
- Baseline model or heuristic, if used.  
- Main model(s) chosen and why they are appropriate.

# 7. Model Implementation & Evaluation

This corresponds to the **Model Implementation & Evaluation** rubric criterion.

Describe:

- Features used.  
- Data splitting strategy (train/test or cross-validation).  
- Metrics used (accuracy, F1, RMSE, etc.).  
- Tables/plots summarizing performance.  
- A short interpretation of each metric/plot.

```{r modeling, eval=FALSE}
# Example structure:
# set.seed(123)
# train_index <- sample(seq_len(nrow(data_clean)), size = 0.8 * nrow(data_clean))
# train <- data_clean[train_index, ]
# test  <- data_clean[-train_index, ]
#
# model <- glm(target ~ ., data = train, family = binomial)
# preds <- predict(model, newdata = test, type = "response")
# # compute metrics...
```

# 8. Conclusions & Recommendations

Summarize the key takeaways:

- Answer your original research question(s) directly.  
- Highlight the most important patterns or relationships you found.  
- Discuss limitations (data size, bias, missing variables, etc.).  
- Suggest possible extensions or future work.

# 9. Code Quality & Reproducibility

Briefly document how someone else can reproduce your results:

- Which R scripts or Rmd files to run.  
- Any required R packages.  

```{r session-info, echo=FALSE}
sessionInfo()
```

# 10. References

List any references you used, such as:

- Dataset documentation.  
- Research papers or articles.  
- Tutorials or blog posts.

https://www.youtube.com/watch?v=gFs1VFviyPs&t=15
https://www.geeksforgeeks.org/r-language/create-heatmap-in-r-using-ggplot2/
https://chatgpt.com/share/693601d8-b3f0-8009-ad35-0e71d9c9034b
https://stackoverflow.com/questions/60086041/how-do-i-use-r-to-count-multiple-variables


# Appendix (Optional)

Include extra plots, diagnostic checks, or model comparisons if needed.
